Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-03-01 grove-paper.tex
move decomp algorithm notes to Implementation
edits with Cyrus and Michael

% \newcommand{\multiVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\curlyveedownarrow_{#1}}}}}
% \newcommand{\cycleVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\rcirclearrowleft_{#1}}}}}

% TODO: make sure we're not claiming to invent the term "unicycular graph"
% TODO: move "algorithmic section" to Implementation section
% TODO: maybe add a forward reference at end here to Implementation section
% TODO: clean up old comments below

@@

The fourth component $U$ is the set of expressions derived from the remaining vertices, which form unicycular graphs where each vertex has exactly one parent.
Every unicycular graph consists of a single cycle where each vertex in the cycle may have additional children that are not part of the cycle.
Breaking the cycle turns a unicyclic graph into a tree.
To break the cycle, we arbitrarily choose the vertex on the cycle with the smallest id according to the total ordering on $\U$.
Formally, we choose the vertex $v$ such that $v = \min(\ancestors(v))$ because
only a vertex on the cycle can be its own ancestor.

@@

% TODO: add conflict hole to no-parent vertex

lemma: a unicycular graph is a cycle with trees hanging off of it

lemma: unicycular graph traversal always produces a valid tree

Unicycular graph traversal occurs in two directions. A vertex is chosen as an arbitrary
starting point and its parents are traversed and marked as seen. Any vertex that
has been seen twice must be on the cycle. We take the first one as the root of
the unicycular graph and traverse its children to produce a tree that covers the unicycular graph.
Any remaining single-parented vertices must be parts of other unicycular graphs and are
traversed until there are no more single-parented vertices left.

@@

% TODO: clean up this figure
% TODO: move to implementation section

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      \textrm{Trees:\qquad}  & t & {}\in Tree & {}::={} & \textrm{Vertex}(v, m) \mid \textrm{Ref}(v) \\
      \textrm{Maps:}         & m & {}\in Map & {}::={} & \varnothing \mid p \mapsto [(\e, t), \ldots, (\e, t)]; m \\
    \end{array}
  \]
  \caption{Syntax of graph decomposition trees as a grammar.}
  \label{fig:Syntax of graph decomposition trees}
\end{figure}

% To render the graph as disjoint fragments of program code, we decompose the
% graph into an equivalent representation using trees. The result is a 4-tuple,
% where the first component is a tree, the second and third components are sets of
% trees, and the fourth component is a set of cycles.

% 1. ($O(n \log n)$ for maps, $O(n)$ for hash table) linear scan to partition vertices according to # of parents: multiple parents, single parents, no parents.

% 2. ($n \log n$?) scan to discover edges leading to single-parented vertices reachable from root (and remove from single-parented set)

% 3. ($n \log n$?) scan to discover edges leading to single-parented vertices reachable from multi-parented vertices (and remove from single-parented set).

% 4. ($n \log n$?) scan to discover edges leading to single-parented [deleted]

% 5. simple cycles—reverse traversal, then forward traversal. Here's where we want to prove some things (e.g., dangling vertices are not deleted)

% tree ::= Vertex (Constructor) | Ref (Vertex.Id) | Conflict (tree list)

%
% ... () of sets of
% (poly)trees) wherein each multi-edge has been resolved into a simple edge with
% one or more symbolic references to it.

% Our decomposition algorithm first separates the graph into three disjoint sets of
% vertices according to the number of parents and presence of simple cycles.
% %
% The first set is the single-parented vertices that have the root as an ancestor.
% The second set is the vertices that have a multi-parented vertex as an ancestor.
% The third

% We then derive equivalent tree representations of connected components by tracing
% the ancestry, taken as the transitive closure of the childhood relation, of the
% vertices in each set.

% set contains the multi-parented vertices and
% their non-multiparented descendants. The second set contains the vertices with
% no parents and their non-multiparented descendants. We can determine both of
% these sets with a single pass over all of the live vertices in the graph. The
% third set contains the single-parented vertices that are descendants of the root
% vertex.

% The first set contains all of the
% vertices that have the root as an ancestor.

% : vertices with multiple parents, vertices with no parents,
% single-parented vertices reachable from the root, and everything else.

% separate it into 
% components that only touch  detect any vertices that either have multiple parents or no parents
% at all.

% partition its vertices into three disjoint sets: multi-parented, deleted,
% and any remaining vertices participating in a live edge. We then partition the
% graph into four non-overlapping sets of trees: multi-parented, single-parented
% reachable from root, deleted, and simple cycles.

%  detect any multi-parented
% vertices and use them to partition the remaining vertices of the graph into .

% Our decomposition algorithm partitions the graph into trees of
% connected components. Cycles are resolved by replacing all but one vertex of
% each multi-edge with a by-id reference.

% Translating a graph into expressions in the calculus allows us to model user
% interaction as a stream of edits as described in \autoref{sub:User Actions}.

% \begin{algorithm}[t]
%   \caption{Graph Decomposition}
%   \label{fig:Graph Decomposition}

%   \DontPrintSemicolon
%   \SetAlgoNoLine
%   \SetKwFunction{reachable}{reachable}
%   \SetKwProg{Fn}{function}{}{end}

%   \KwIn{Graph $G$}
%   \KwOut{A partitioning of the edges of $G$}

%   $live := \{\e : G(\e) = \Plus\}$\;
%   $R := \{\reachable(\e) : \e = (u, v_0, p, v) \in live\}$\;
%   $D := \{\reachable(\e) : \e \notin live\}$\;
%   $muliparent := \{(u,v,p,v') \in live : \exists (u_1, v_1, p_1, v), (u_2, v_2, p_2, v) \in live, v_1 \ne v_2\}$\;
%   $MP := \{\reachable(\e) : \e \in multiparent\}$\;
%   $SC := \varnothing$\;
%   $remaining := \{\e \notin \edges(R \cup D \cup MP)\}$\;
%   \While{$remaining \ne \varnothing$}{
%     $(root, cycle) := \reachable(\min(remaining))$\;
%     $remaining := remaining \setminus \{root\} \setminus cycle$\;
%     $SC := SC \cup \{(root, cycle)\}$\;
%   }
%   \Return{$(R, D, MP, SC)$}\;
%   \;

%   \Fn{$\reachable\ ((u, v, p, v'))$}{
%     $children := \{(u', v', p', v'') \in live\}$\;
%     $descendants := \bigcup_{\e \in children} \reachable(\e)$\;
%     \Return{$((u, v, p, v'), descendants)$}
%   }
% \end{algorithm}

% The result of decomposition is a 4-tuple $(e, MP, D, SC) \in Exp \times
% \wp({Exp}) \times \wp({Exp}) \times \wp({Exp})$. Expression $e$ covers the
% portion of the graph reachable from the root vertex $v_0$. $MP$ contains the
% expressions that involve vertices with multiple parents. $D$ contains deleted
% expressions, i.e., those not reachable from $v_0$, that do not contain cycles.
% $SC$ contains the deleted expressions that do not involve multiple parents but
% do contain cycles.

% A simple cycle is a set of vertices connected by a closed path, consisting
% of:

% 1) two or more vertices, all descendants of each other, and
% 2) the vertices of any non-multiparent subtrees reachable from the core, and
% not reachable from the root vertex.

% To determine the simple cycles of a graph, we:

% 1) check if any vertex in the graph is a descendant of itself or some other
% vertex that is (a descendant of itself),
% 2) find the least vertex in the core of this new(ly discovered) cycle, and
% 3) return a tree connecting every vertex reachable from the least (vertex).

% The least vertex of a cycle is the one that was created first. To ensure consistent behavior, the search for reachable vertices begins at the least vertex in a cycle.

% <Mathematical definition of above.>

@@

@@

>>>>

% TODO: make sure we're not claiming to invent the term "unicycular graph"
% TODO: maybe add a forward reference at end here to Implementation section
% TODO: make ID of y lower than the unicycle root, to show what we're minimizing

@@

The fourth component $U$ is the set of expressions derived from the remaining vertices which necessarily have exactly one parent.
These vertices form structures formally called unicycular graphs~\citep{DBLP:journals/algorithmica/KruskalRS90}, or \emph{unicycles} for short.
Every unicycle consists of a single cycle where each vertex in the cycle may have additional children that are not part of the cycle.
Breaking the cycle turns a unicyclic graph into a tree.
% TODO: try clearing up "the root" as e.g., "the component root" or "the root of this components of the grove"
% TODO: make sure we say "the graph root" when talking about the actual root, and "the component root" everywhere else.
To break the cycle, we arbitrarily choose the vertex on the cycle with the smallest id, $u$, as the root of that tree.
Formally, we choose the vertex $v$ such that $v = \min(\ancestors(v))$ because
only a vertex on the cycle can be its own ancestor.
When an edge to this vertex appears, we place a reference to it $\cycleVertex{u}$.
In our example, ${+}_{\id{46}}$ is referenced by the $LEFT$ edge of ${*}_{\id{48}}$.

@@

@@

% TODO: maybe add a forward reference here to Implementation section

@@

% TODO: talk about code generation somewhere in here

% TODO: talk about:
% - decomp vertex set generation and unicycle traversal
% - how to choose which branch of expr' to take
% - lfp(ancestors') and min of it

% TODO: can we save any of this in a README or code comments?

@@

% TODO: use code generation instead: have Lang.ml generate expr / expr' / ... for us; then we don't need this figure.
% TODO: maybe show this figure and say that it's possible to handle generically without code generation

% \begin{figure}
%   \[
%     \arraycolsep=0pt
%     \begin{array}{lrlll}
%       \textrm{Trees:\qquad}  & t & {}\in Tree & {}::={} & \textrm{Vertex}(v, m) \mid \textrm{Ref}(v) \\
%       \textrm{Maps:}         & m & {}\in Map & {}::={} & \varnothing \mid p \mapsto [(\e, t), \ldots, (\e, t)]; m \\
%     \end{array}
%   \]
%   \caption{Syntax of graph decomposition trees as a grammar.}
%   \label{fig:Syntax of graph decomposition trees}
% \end{figure}

lemma: a unicycle is a cycle with trees hanging off of it

lemma: unicycle traversal always produces a valid tree

Unicycular graph traversal occurs in two directions. A vertex is chosen as an arbitrary
starting point and its parents are traversed and marked as seen. Any vertex that
has been seen twice must be on the cycle. We take the first one as the root of
the unicycle and traverse its children to produce a tree that covers the unicycular graph.
Any remaining single-parented vertices must be parts of other unicycular graphs and are
unicycle-traversed until there are no more single-parented vertices left.
